Current state and prospects of R-packages for the design of experiments

### 2022-12-13 paper.Rmd
huge update for index with cleaner pipeline

---
title: Current state and prospects of R-packages for the design of experiments
author:
  - name: Emi Tanaka
    affiliation: Monash University
  - name: Dewi Amaliah 
    affiliation: Monash University
abstract: >
    Re-running an experiment is generally costly and in some cases impossible due to limited resources, so the design of an experiment plays a critical role in increasing the quality of experimental data.  In this paper we describe the current state of the R-packages for the design of experiments through an exploratory data analysis of package downloads, package metadata, and the comparison of characteristics with other topics. We observe that experimental designs in practice appear to be sufficiently manufactured by a small number of packages and the development of experimental designs often occur in silos. We discuss also the interface designs of widely utilised R packages in the field of experimental design and discuss its future prospects to advance the field in practice.
bibliography: paper.bib
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(inputFile), 'index.html')) })
keywords: ["experimental design", "CRAN task view", "interface design"]
output: 
  pagedown::html_paged:
    template: assets/html-template.html
    csl: assets/html-template.csl
    self_contained: false
    css: ["assets/html-template.css"]
---

```{r setup, cache = FALSE, include = FALSE}
library(targets)
library(glue)
library(ggwordcloud)
library(tidyverse)
library(patchwork)
library(tidygraph)
library(ggraph)
source("R/plot.R")
knitr::opts_chunk$set(echo = FALSE, 
                      cache = TRUE,
                      cache.path = "cache/",
                      fig.align = 'center', 
                      fig.pos = 'htbp', 
                      fig.path = "figures/",
                      fig.width = 6,
                      fig.height = 6.5,
                      message = FALSE,
                      warning = FALSE)

ctv <- function(view) {
  ifelse(knitr::is_html_output(), glue("[{view}](https://cran.r-project.org/web/views/{view}.html)"), glue("\\ctv{[view]}", .open = "[", .close = "]"))
}
CRANpkg <- function(pkg) {
  ifelse(knitr::is_html_output(), glue("[{pkg}](https://cran.r-project.org/web/packages/{pkg}/index.html)"), glue("\\CRANpkg{[pkg]}", .open = "[", .close = "]"))
}
ref <- function(x) {
  ifelse(knitr::is_html_output(), glue("\\@ref({x})"), glue("\\ref{[x]}", .open = "[", .close = "]"))
}
url <- function(x) {
  ifelse(knitr::is_html_output(), x, glue("\\url{[x]}", .open = "[", .close = "]"))
}

tar_load(npkgs)
tar_load(gini_yearly_doe)
tar_load(gini_yearly_ctv)
tar_load(doe_lorenz2021)

lorenz_value <- function(p) {
  doe_lorenz2021 %>% 
    mutate(pdist = abs(p - .env$p)) %>% 
    filter(pdist == min(pdist)) %>% 
    pull(L) %>% 
    magrittr::multiply_by(100) %>% 
    round()
}
```

@@

<!-- Unlike many other data collection schemes, experiments provides the greatest control and consequently, a greater opportunity to statistically optimise the information drawn from the experimental data. -->

@@

In `r ctv("ExperimentalDesign")` task view, there are `r npkgs` R packages for experimental design and analysis of data from experiments. The sheer quantity and variation of experimental designs in the R-packages are arguably unmatched with any other programming languages, e.g. in Python, only a handful of packages that generate design of experiment exist (namely `pyDOE`, `pyDOE2`, `dexpy`, `experimenter` and `GPdoemd`) with limited type of designs. Thus, the study of DoE packages, based on quantitative and qualitative data, can give us an objective view to the state of the current experimental designs in practice.

A utility of software can also be described by its design to facilitate clear expression and interpretation of the desired experimental design. Certain programming language design can hinder or discourage development of a reliable program [@Wasserman1975-xr].  The immense popularity of `tidyverse` [a collection of R-packages for various stages of data analysis that place enormous emphasis on the interface design by @Wickham2019-mj] is a testament to the impact an interface design can have in practice. The practice of experimental design could be advanced by adopting similar interface design principles across DoE packages.

@@

The Comprehensive R Archive Network (CRAN) is a network of servers located across the world that store mirrored versions of R and R-packages. The most popular network is the RStudio mirror (the default server for those that use the RStudio IDE). The RStudio mirror is also the only server that provides a comprehensive daily download logs of R and R-packages since October 2012. The summary data can be easily accessed with the `cranlogs` package [@cranlogs]. This paper uses the data from the beginning of 2013 to end of 2021 (a total of 9 years) for the packages in the CRAN task views.

@@

All CRAN packages have a title, description, package connections (suggests, depends and imports of other packages) and other meta-information in the DESCRIPTION file. We use the text data from the title and description (as accessed in `r as.Date(file.info("_targets/objects/ngram_download_data")$mtime)`) in Section `r ref("topics")`.

@@

CRAN task views are volunteer maintained list of R-packages on CRAN relevant to the corresponding topic. There are a total `r n_distinct(gini_yearly_ctv$ctv)` CRAN task views. Table S1 in the Supplementary Materials list the full available topics from the `ctv` package [@ctv].
The list of packages in each CRAN task views (as of `r as.Date(file.info("_targets/objects/ctv_pkgs_data")$mtime)`) are used to contrast the characteristics of DoE packages in Sections `r ref("popular")` and `r ref("silo")`.

@@

There are at least 50 DoE packages since 2013 but most of the downloads are concentrated in just a handful of packages. For example, Figure `r ref("fig:plot-lorenz")` shows a Lorenz curve [@Lorenz1905-tc] for the total package downloads in 2021 of `r tar_read(gini_yearly) %>% filter(year==2021) %>% pull(n)` DoE packages (first released prior to 2021); we can see from Figure `r ref("fig:plot-lorenz")` that bottom 90\% of DoE packages (in terms of total download count in 2021) only share about `r lorenz_value(0.9)`\% of total downloads across all DoE packages -- in another words, `r 100 - lorenz_value(0.9)`\% of the total downloads are due to `r round(npkgs * 0.1)` packages (10% of the DoE packages).

```{r plot-lorenz, fig.height = 4, fig.width = 4, fig.cap = "Lorenz curve of the total download count for DoE packages in 2021. The red line corresponds to the line of perfect equality. The yellow region shows the area under the Lorenz curve and the red region shows the area of gap in equality."}
tar_read(plot_lorenz2021) +
  plot_setup() +
  labs(y = "Cumulative share of downloads (%)",
       x = "Percentage of DoE packages") +
  theme(plot.margin = margin(r = 20, t = 20)) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  scale_y_continuous(breaks = seq(0, 100, by = 10))
```

 If we consider package downloads as a measure of "wealth", then we can consider using the Gini index [@Gini1921-mf] as a measure of download inequality across packages. The ratio of the red region over the total colored regions in Figure `r ref("fig:plot-lorenz")` corresponds to the Gini index for 2021. A Gini index of 0\% indicates equality in downloads across packages while a value of 100\% indicates maximal inequality (all downloads are due to one package). In Figure `r ref("fig:download-share")`, we see that the distributions of the package downloads each year have a heavy right tail with the Gini index ranging from `r scales::percent(min(gini_yearly_doe$gini), 0.1)` to `r scales::percent(max(gini_yearly_doe$gini), 0.1)` across the years 2013 to 2021, indicating that there is a high level of inequality of package downloads, particularly with more pronounced inequality in the last 6 years.

```{r download-share, fig.width = 5.5, fig.height = 4, fig.cap = "The distribution of the number of downloads for DoE packages by year. Packages were removed in any year if it was released in that year or later so that each download count is for the full year. The label on the bottom of the plot shows the Gini index for downloads and the number of packages with full year of download count in the corresponding year. In the last 6 years, the Gini index is consistent above 60\\% indicating that most downloads are due to a relatively small number of packages. "}
tar_read(plot_download_dist) +
  labs(y = "Number of downloads")
```

The increase in the number of packages that are not highly downloaded may mean that **_there are more packages to construct niche experimental designs_**. Some examples of these packages include `qtlDesign`, `PwrGSD` and `Crossover` made for QTL experiments, group sequential designs and crossover trials, respectively. These packages would naturally have a smaller number of potential users. Counterfactual to this, the increase could be due to other external factors, such as an increase in the number of skilled developers (and thus more package contributions), a change in CRAN policy or management to add packages (either to CRAN and/or task view), and/or that new packages are still yet to amass users. While there is an argument that low download counts are due to the low utility and/or quality of the packages, packages in CRAN task views are selected by expert maintainers, thus we can reasonably assume that any package listed in CRAN task view are of a decent utility and quality.

@@

While in absolute terms the Gini index is high for `r ctv("ExperimentalDesign")` task view (`r scales::percent(min(gini_yearly_doe$gini), 0.1)` to `r scales::percent(max(gini_yearly_doe$gini), 0.1)`), the inequality is not as severe as other CRAN task views as shown in Figure `r ref("fig:fig-gini-all-ctvs")`. We can see in Figure `r ref("fig:fig-gini-all-ctvs")` that the Gini index is generally increasing for DoE packages (as is generally the case for other CRAN task views as shown in Figure S1 in the Supplementary Materials) but most other CRAN task views have a Gini index of over 75%. This suggests that other CRAN task views may have dominant standards and in comparison to other topics, there are more **_diverse approaches to designing experiments_** and thus no single DoE package is dominant. This observation however doesn't take into account other approaches to generate experimental designs, like say the proprietary software, CycDesignN [@cycdesignn], which may be widely used.

```{r fig-gini-all-ctvs, fig.height = 4, fig.width = 4, fig.cap = "The points show the Gini index of the download counts by year for ExperimentalDesign task view with the color showing the number of packages. The grey lines show the line plots of the Gini index across years for all other CRAN task views. See Figure S1 in the Supplementary Material for the line graph for the Gini index across years by each CRAN task view."}
gini_yearly_ctv %>% 
  ungroup() %>%  
  arrange(ctv, year) %>% 
  mutate(ctv = fct_reorder(ctv, gini, last),
         year = factor(year),
         gini = 100 * gini) %>%  
  ggplot(aes(year, gini, group = ctv)) + 
  geom_line(color = "grey") +
  geom_line(data = ~filter(.x, ctv=="ExperimentalDesign")) + 
  geom_point(data = ~filter(.x, ctv=="ExperimentalDesign"),
             aes(color = n))  +
  plot_setup() +
  labs(x = "Year", y = "Gini index (%)", color = "# of packages",
       title = "ExperimentalDesign CRAN task view") +
  guides(color = guide_colorbar(direction = "horizontal")) +
  theme(legend.position = "bottom") +
  scale_colour_distiller(palette = "YlOrRd", direction = 1) 
```

@@

We can see in Figure `r ref("fig:rank-over-time")` that most of the top 10 ranking packages have been in the top 10 for the last 9 years with `lhs` steadily climbing up the ranks in the last few years. It should be noted that the download of one package can prompt a download of another package; the most notable package connection is `AlgDesign` and `agricolae`, where the former is an import for the latter. The full network of packages connections within the DoE packages is shown later in Figure `r ref("fig:plot-doe-network")`.

```{r rank-over-time, fig.width = 5.5, fig.height = 4.5, fig.cap = "The plot shows the rank of top 10 downloaded packages by year. Packages that do not appear in top 10 for at least two periods are omitted from the plot. Most packages are consistently in the top 10 for the period shown."}
tar_read(plot_rank_dist) + labs(color = "Package")
```

@@

```{r release-date-vs-download, fig.height = 5, fig.cap = "The above figure shows the total download (in log scale) of a package in the corresponding year against the first release date of the package. The blue line corresponds to the least squares fit of a simple linear regression model. The label in the right-hand upper corner shows the sample correlation coefficient between the first release date and the log (with base 10) of the total download count. The high-leverage point on the far left belongs to `conf.design`, authored by one of the earlier contributors to R."}
tar_read(cran_rank_yearly) %>%
  mutate(nupdates = map2_int(package, year, ~{
    tar_read(updates_all) %>% 
      filter(package == .x,
             lubridate::year(update) <= .y) %>% 
      nrow()
  })) %>% 
  ggplot(aes(first, total/1000)) +
  geom_point(aes(color = nupdates)) + 
  geom_label(data = ~ .x %>% 
              group_by(year) %>% 
              summarise(R = cor(as.numeric(first), log10(total/1000)) %>% 
                          scales::comma(0.01) %>% 
                          paste0("r = ", .),
                        y = quantile(total/1000, 0.99)),
            aes(label = R, y = y, x = as.Date("2014-11-01")),
            size = 3) +
  facet_wrap(~year, scale = "free_y") +
  geom_smooth(se = FALSE, 
              method = lm, 
              formula = y ~ x) +
  scale_y_log10() + 
  scale_color_continuous(trans = "log10") + 
  plot_setup() +
  labs(x = "First release date",
       y = "Total download (1000s)",
       color = "# of updates") +
  theme(legend.position = "bottom")
```

@@

1. We standardised the words to lower case and removed pluralization. 
2. Multiple mentions of the same bigram within a package was counted as one (e.g. `AlgDesign` mentions "experimental design" four times in the title and the description but this is counted as one). 
3. Bigrams that consist the stop words are removed. The stop words are sourced from the lexicons in `tidytext::stop_words` in addition to other words we deemed irrelevant, e.g. "provide", "e.g.", "calculate" and so on -- full list is shown in the code provided in the link under Section `r ref("pkgs")`.

@@

```{r wordcloud-over-time, fig.width = 5.5, fig.height = 5, fig.cap = "The above figure shows the word cloud of bigrams from the title and descriptions of the DoE packages. The size shows how often the bigram appears across the DoE packages and the color are relative to the total download count in 2021 of the packages that contain the bigram."}
tar_read(plot_wordcloud2) + 
  scale_size_area(max_size = 5) +
  theme(legend.text = element_text(size = 10),
        legend.title = element_text(size = 12)) +
  labs(color = "Total downloads (%)")
```

@@

Figure `r ref("fig:ctv-summ-plot")` shows that the `r ctv("ExperimentalDesign")` task view has the lowest average number of contributors among all `r n_distinct(gini_yearly_ctv$ctv)` CRAN task views. In addition, we can also see in Figure `r ref("fig:ctv-summ-plot")`  that the `r ctv("ExperimentalDesign")` task view has one of the least intra-connectivity (the percentage of packages that make use of other packages within the same task view). The full connection of DoE packages with other DoE packages is shown in Figure `r ref("fig:plot-doe-network")`. These observations suggest that the **_field of experimental design is one of the least collaborative field_** and package development generally occur in silos.

```{r ctv-summ-plot, fig.width = 5.5, fig.height = 6, fig.cap = "The above figure is a scatterplot of the intra-connectivity (the percentage of packages that depends, suggest or imports at least one other package within the same task view) and the average number of contributors for each CRAN task view. A low intra-connectivity suggests that development within the topic mostly occur in silos whilst high  intra-connectivity suggests that there are more interactions within the topic. The color shows the number of packages, the size of the point corresponds to the total number of contributors, and the text labels show the CRAN task view name.  The label of ExperimentalDesign task view is colored red. The task views in the bottom-left corner are topics that are more indicative of contributors working in silos. The actual numerical values are show in Table S1 in the Supplementary Material."}
tar_read(ctv_summary_table) %>% 
  ggplot(aes(avg_contributors, intra_perc)) +
  geom_point(aes(size = ncontributors,
                 color = n)) +
  ggnewscale::new_scale_color() + 
  ggrepel::geom_text_repel(aes(label = name,
                               color = name=="ExperimentalDesign"),
                           force = 150,
                           max.iter = 20000) +
  guides(color = "none") + 
  scale_color_manual(values = c("black", "#CC3311")) + 
  labs(x = "Average number of contributors",
       y = "Intra-connectivity (%)",
       size = "Total number of contributors",
       color_new = "Total number of packages") +
  plot_setup() +
  theme(legend.position = "bottom",
        legend.box = "vertical",
        legend.box.just = "left")
```

>>>>

---
title: Current state and prospects of R-packages for the design of experiments
author:
  - name: Emi Tanaka
    affiliation: Monash University
  - name: Dewi Amaliah 
    affiliation: Monash University
abstract: >
    Re-running an experiment is generally costly and in some cases impossible due to limited resources, so the design of an experiment plays a critical role in increasing the quality of experimental data.  In this paper we describe the current state of the R-packages for the design of experiments through an exploratory data analysis of package downloads, package metadata, and the comparison of characteristics with other topics. We observe that experimental designs in practice appear to be sufficiently manufactured by a small number of packages and the development of experimental designs often occur in silos. We discuss also the interface designs of widely utilised R packages in the field of experimental design and discuss its future prospects to advance the field in practice.
bibliography: paper.bib
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(inputFile), 'index.html')) })
keywords: ["experimental design", "CRAN task view", "interface design"]
output: 
  pagedown::html_paged:
    template: assets/html-template.html
    csl: assets/html-template.csl
    self_contained: true
    css: ["assets/html-template.css"]
---

```{r setup, cache = FALSE, include = FALSE}
library(glue)
library(ggwordcloud)
library(tidyverse)
library(patchwork)
library(tidygraph)
library(ggraph)
theme_set(
  theme(
    panel.background = element_rect(fill = NA),
    panel.grid = element_line(color = "lightgray"),
    axis.text = element_text(color = "black"),
    axis.line = element_line(color = "black", size = 0.7),
    axis.ticks.length = unit(1.4, "mm"),
    axis.ticks = element_line(color = "black", size = 0.7),
    axis.title = element_text(color = "black", face = "bold"),
    strip.background = element_rect(color = "black",
                                    fill = "white"),
    strip.text = element_text(color = "black",
                              margin = margin(5, 5, 5, 5)),
    plot.title.position = "plot",
    plot.title = element_text(color = "black", hjust = 0)
  )
)
knitr::opts_chunk$set(echo = FALSE, 
                      cache = TRUE,
                      cache.path = "cache/",
                      fig.align = 'center', 
                      fig.pos = 'htbp', 
                      fig.path = "figures/",
                      fig.width = 6,
                      fig.height = 6.5,
                      message = FALSE,
                      warning = FALSE)

ctv <- function(view) {
  ifelse(knitr::is_html_output(), glue("[{view}](https://cran.r-project.org/web/views/{view}.html)"), glue("\\ctv{[view]}", .open = "[", .close = "]"))
}
CRANpkg <- function(pkg) {
  ifelse(knitr::is_html_output(), glue("[{pkg}](https://cran.r-project.org/web/packages/{pkg}/index.html)"), glue("\\CRANpkg{[pkg]}", .open = "[", .close = "]"))
}
ref <- function(x) {
  case_when(knitr::is_html_output() ~ glue("\\@ref({x})"),
            TRUE ~ glue("\\ref{[x]}", .open = "[", .close = "]"))
}
url <- function(x) {
  ifelse(knitr::is_html_output(), x, glue("\\url{[x]}", .open = "[", .close = "]"))
}
rpkg <- function(x) {
  glue("`{x}`")
}

ctv_pkg_data <- readRDS("output/ctv_pkg_data.rds")
ctv_summary_table <- readRDS("output/ctv_summary_table.rds")
doe_bigram_all_with_download_data <- readRDS("output/doe_bigram_all_with_download_data.rds")
doe_pkg_network <- readRDS("output/doe_pkg_network.rds")
ctv_gini_yearly <- readRDS("output/ctv_gini_yearly.rds")
ctv_lorenz_yearly <- readRDS("output/ctv_lorenz_yearly.rds")
ctv_rank_yearly <- readRDS("output/ctv_rank_yearly.rds")
access_date_pkg_downloads <- readRDS("output/access_date_pkg_downloads.rds")

doe_npkgs <- ctv_summary_table %>% 
  filter(name=="ExperimentalDesign") %>% 
  pull(n)

doe_npkgs_2021 <- ctv_gini_yearly %>% 
  filter(year==2021, name=="ExperimentalDesign") %>% 
  pull(n)

doe_ginis <- ctv_gini_yearly %>% 
  filter(name == "ExperimentalDesign") %>% 
  pull(gini)

lorenz_value <- function(p) {
  ctv_lorenz_yearly %>% 
    filter(name == "ExperimentalDesign",
           year == 2021) %>% 
    unpack(lorenz_data) %>% 
    mutate(pdist = abs(p - .env$p)) %>% 
    filter(pdist == min(pdist)) %>% 
    pull(L) %>% 
    magrittr::multiply_by(100) %>% 
    round()
}
```

@@

@@

In `r ctv("ExperimentalDesign")` task view, there are `r doe_npkgs` R packages for experimental design and analysis of data from experiments. The sheer quantity and variation of experimental designs in the R-packages are arguably unmatched with any other programming languages, e.g. in Python, only a handful of packages that generate design of experiment exist (namely `pyDOE`, `pyDOE2`, `dexpy`, `experimenter` and `GPdoemd`) with limited type of designs. Thus, the study of DoE packages, based on quantitative and qualitative data, can give us an objective view to the state of the current experimental designs in practice.

A utility of software can also be described by its design to facilitate clear expression and interpretation of the desired experimental design. Certain programming language design can hinder or discourage development of a reliable program [@Wasserman1975-xr].  The immense popularity of `r rpkg("tidyverse")` [a collection of R-packages for various stages of data analysis that place enormous emphasis on the interface design by @Wickham2019-mj] is a testament to the impact an interface design can have in practice. The practice of experimental design could be advanced by adopting similar interface design principles across DoE packages.

@@

The Comprehensive R Archive Network (CRAN) is a network of servers located across the world that store mirrored versions of R and R-packages. The most popular network is the RStudio mirror (the default server for those that use the RStudio IDE). The RStudio mirror is also the only server that provides a comprehensive daily download logs of R and R-packages since October 2012. The summary data can be easily accessed with the `r rpkg("cranlogs")` package [@cranlogs]. This paper uses the data from the beginning of 2013 to end of 2021 (a total of 9 years) for the packages in the CRAN task views.

@@

All CRAN packages have a title, description, package connections (suggests, depends and imports of other packages) and other meta-information in the DESCRIPTION file. We use the text data from the title and description (as accessed in `r access_date_pkg_downloads`) in Section `r ref("topics")`.

@@

CRAN task views are volunteer maintained list of R-packages on CRAN relevant to the corresponding topic. There are a total `r n_distinct(ctv_pkg_data$name)` CRAN task views. Table S1 in the Supplementary Materials list the full available topics from the `r rpkg("ctv")` package [@ctv].
The list of packages in each CRAN task views (as of `r access_date_pkg_downloads`) are used to contrast the characteristics of DoE packages in Sections `r ref("popular")` and `r ref("silo")`.

@@

There are at least 50 DoE packages since 2013 but most of the downloads are concentrated in just a handful of packages. For example, Figure `r ref("fig:plot-lorenz")` shows a Lorenz curve [@Lorenz1905-tc] for the total package downloads in 2021 of `r doe_npkgs_2021` DoE packages (first released prior to 2021); we can see from Figure `r ref("fig:plot-lorenz")` that bottom 90\% of DoE packages (in terms of total download count in 2021) only share about `r lorenz_value(0.9)`\% of total downloads across all DoE packages -- in another words, `r 100 - lorenz_value(0.9)`\% of the total downloads are due to `r round(doe_npkgs_2021 * 0.1)` packages (10% of the DoE packages).

```{r plot-lorenz, fig.height = 4, fig.width = 4, fig.cap = "Lorenz curve of the total download count for DoE packages in 2021. The red line corresponds to the line of perfect equality. The yellow region shows the area under the Lorenz curve and the red region shows the area of gap in equality."}
ctv_lorenz_yearly %>% 
  filter(name == "ExperimentalDesign",
         year == 2021) %>% 
  ggplot(aes(lorenz_data$p * 100, lorenz_data$L * 100)) +
  geom_abline(color = "red") +
  geom_ribbon(aes(ymin = 0, ymax = lorenz_data$L * 100),
                fill = "#FFC107", alpha = 0.5) +
  geom_ribbon(aes(ymin = lorenz_data$L * 100, ymax = lorenz_data$p * 100),
                 fill = "#D81B60", alpha = 0.5) +
  geom_line() + 
  geom_point() +
  annotate("text", x = 75, y = 80, 
           label = "Line of perfect equality",
            angle = 45) +
  labs(y = "Cumulative share of downloads (%)",
       x = "Percentage of DoE packages") +
  theme(plot.margin = margin(r = 20, t = 20)) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  scale_y_continuous(breaks = seq(0, 100, by = 10))
```

 If we consider package downloads as a measure of "wealth", then we can consider using the Gini index [@Gini1921-mf] as a measure of download inequality across packages. The ratio of the red region over the total colored regions in Figure `r ref("fig:plot-lorenz")` corresponds to the Gini index for 2021. A Gini index of 0\% indicates equality in downloads across packages while a value of 100\% indicates maximal inequality (all downloads are due to one package). In Figure `r ref("fig:download-share")`, we see that the distributions of the package downloads each year have a heavy right tail with the Gini index ranging from `r scales::percent(min(doe_ginis), 0.1)` to `r scales::percent(max(doe_ginis), 0.1)` across the years 2013 to 2021, indicating that there is a high level of inequality of package downloads, particularly with more pronounced inequality in the last 6 years.

```{r download-share, fig.width = 5.5, fig.height = 4, fig.cap = "The distribution of the number of downloads for DoE packages by year. Packages were removed in any year if it was released in that year or later so that each download count is for the full year. The label on the bottom of the plot shows the Gini index for downloads and the number of packages with full year of download count in the corresponding year. In the last 6 years, the Gini index is consistent above 60\\% indicating that most downloads are due to a relatively small number of packages. "}
ctv_rank_yearly %>% 
    filter(name == "ExperimentalDesign") %>% 
    mutate(year = as.factor(year)) %>% 
    ggplot(aes(x = year, y = total)) + 
    geom_violin() +
    geom_boxplot(width = 0.1) + 
    labs(x = "Year", y = "Number of downloads") + 
    scale_y_log10(label = scales::comma, expand = c(0.1, 0)) +
    geom_label(data = ctv_gini_yearly %>% 
                        filter(name == "ExperimentalDesign") %>% 
                        mutate(year = as.factor(year)), 
               size = 3,
               aes(y = 900, 
                   label = paste0(scales::percent(gini, 0.1),
                                   "\nn=", n)))
```

The increase in the number of packages that are not highly downloaded may mean that **_there are more packages to construct niche experimental designs_**. Some examples of these packages include `r rpkg("qtlDesign")`, `r rpkg("PwrGSD")` and `r rpkg("Crossover")` made for QTL experiments, group sequential designs and crossover trials, respectively. These packages would naturally have a smaller number of potential users. Counterfactual to this, the increase could be due to other external factors, such as an increase in the number of skilled developers (and thus more package contributions), a change in CRAN policy or management to add packages (either to CRAN and/or task view), and/or that new packages are still yet to amass users. While there is an argument that low download counts are due to the low utility and/or quality of the packages, packages in CRAN task views are selected by expert maintainers, thus we can reasonably assume that any package listed in CRAN task view are of a decent utility and quality.

@@

While in absolute terms the Gini index is high for `r ctv("ExperimentalDesign")` task view (`r scales::percent(min(doe_ginis), 0.1)` to `r scales::percent(max(doe_ginis), 0.1)`), the inequality is not as severe as other CRAN task views as shown in Figure `r ref("fig:fig-gini-all-ctvs")`. We can see in Figure `r ref("fig:fig-gini-all-ctvs")` that the Gini index is generally increasing for DoE packages (as is generally the case for other CRAN task views as shown in Figure S1 in the Supplementary Materials) but most other CRAN task views have a Gini index of over 75%. This suggests that other CRAN task views may have dominant standards and in comparison to other topics, there are more **_diverse approaches to designing experiments_** and thus no single DoE package is dominant. This observation however doesn't take into account other approaches to generate experimental designs, like say the proprietary software, CycDesignN [@cycdesignn], which may be widely used.

```{r fig-gini-all-ctvs, fig.height = 4, fig.width = 4, fig.cap = "The points show the Gini index of the download counts by year for ExperimentalDesign task view with the color showing the number of packages. The grey lines show the line plots of the Gini index across years for all other CRAN task views. See Figure S1 in the Supplementary Material for the line graph for the Gini index across years by each CRAN task view."}
ctv_gini_yearly %>% 
  arrange(name, year) %>% 
  mutate(name = fct_reorder(name, gini, last),
         year = factor(year),
         gini = 100 * gini) %>%  
  ggplot(aes(year, gini, group = name)) + 
  geom_line(color = "grey") +
  geom_line(data = ~filter(.x, name=="ExperimentalDesign")) + 
  geom_point(data = ~filter(.x, name=="ExperimentalDesign"),
             aes(color = n)) +
  labs(x = "Year", y = "Gini index (%)", color = "# of packages",
       title = "ExperimentalDesign CRAN task view") +
  guides(color = guide_colorbar(direction = "horizontal")) +
  theme(legend.position = "bottom") +
  scale_colour_distiller(palette = "YlOrRd", direction = 1) 
```

@@

We can see in Figure `r ref("fig:rank-over-time")` that most of the top 10 ranking packages have been in the top 10 for the last 9 years with `r rpkg("lhs")` steadily climbing up the ranks in the last few years. It should be noted that the download of one package can prompt a download of another package; the most notable package connection is `r rpkg("AlgDesign")` and `r rpkg("agricolae")`, where the former is an import for the latter. The full network of packages connections within the DoE packages is shown later in Figure `r ref("fig:plot-doe-network")`.

```{r rank-over-time, fig.width = 5.5, fig.height = 4.5, fig.cap = "The plot shows the rank of top 10 downloaded packages by year. Packages that do not appear in top 10 for at least two periods are omitted from the plot. Most packages are consistently in the top 10 for the period shown."}
nrank <- 10
pkgs_repeat_toprank <- ctv_rank_yearly %>% 
  filter(name=="ExperimentalDesign") %>% 
  filter(rank <= nrank) %>% 
  count(package) %>% 
  filter(n > 1) %>% 
  pull(package)

ctv_rank_yearly %>% 
  filter(name=="ExperimentalDesign",
         package %in% pkgs_repeat_toprank) %>% 
  mutate(year = as.factor(year)) %>% 
  ggplot(aes(year, rank, color = package)) +
  geom_point() + 
  geom_line(aes(group = package)) + 
  scale_y_reverse(breaks = 1:nrank) +
  geom_text(aes(label = package),
            data = ~filter(.x, year == 2021),
            hjust = 0,
            nudge_x = 0.1) +
  guides(color = "none") + 
  scale_x_discrete(expand = expansion(c(0.05, 0.25))) +
  labs(color = "Package", x = "Year", y = "Download rank") +
  scale_color_manual(values = c("#E69F00", "#56B4E9", "#009E73", 
                                  "#F0E442", "#0072B2", "#D55E00", 
                                  "#CC79A7", "#000000", "#888888",
                                  "#88CCEE", "#CC6677", "seagreen4", 
                                  "violetred4"))
```

@@

```{r release-date-vs-download, fig.height = 5, fig.cap = "The above figure shows the total download (in log scale) of a package in the corresponding year against the first release date of the package. The blue line corresponds to the least squares fit of a simple linear regression model. The label in the right-hand upper corner shows the sample correlation coefficient between the first release date and the log (with base 10) of the total download count. The high-leverage point on the far left belongs to `conf.design`, authored by one of the earlier contributors to R."}
ctv_rank_yearly %>% 
  filter(name == "ExperimentalDesign") %>% 
  ggplot(aes(first, total/1000)) +
  geom_point(aes(color = nupdates)) + 
  geom_label(data = ~ .x %>% 
              group_by(year) %>% 
              summarise(R = cor(as.numeric(first), log10(total/1000)) %>% 
                          scales::comma(0.01) %>% 
                          paste0("r = ", .),
                        y = quantile(total/1000, 0.99)),
            aes(label = R, y = y, x = as.Date("2014-11-01")),
            size = 3) +
  facet_wrap(~year, scale = "free_y") +
  geom_smooth(se = FALSE, 
              method = lm, 
              formula = y ~ x) +
  scale_y_log10() + 
  scale_color_continuous(trans = "log10") + 
  labs(x = "First release date",
       y = "Total download (1000s)",
       color = "# of updates") +
  theme(legend.position = "bottom")
```

@@

1. We standardised the words to lower case and removed pluralization. 
2. Multiple mentions of the same bigram within a package was counted as one (e.g. `r rpkg("AlgDesign")` mentions "experimental design" four times in the title and the description but this is counted as one). 
3. Bigrams that consist the stop words are removed. The stop words are sourced from the lexicons in `tidytext::stop_words` in addition to other words we deemed irrelevant, e.g. "provide", "e.g.", "calculate" and so on -- full list is shown in the code provided in the link under Section `r ref("pkgs")`.

@@

```{r wordcloud-over-time, fig.width = 5.5, fig.height = 5, fig.cap = "The above figure shows the word cloud of bigrams from the title and descriptions of the DoE packages. The size shows how often the bigram appears across the DoE packages and the color are relative to the total download count in 2021 of the packages that contain the bigram."}
doe_bigram_all_with_download_data %>%
  filter(perc >= 1, n > 1, year == 2021) %>%
  ggplot(aes(
    label = word,
    color = perc,
    size = n
  )) +
  geom_text_wordcloud(shape = "square",
                      show.legend = TRUE) +
  guides(size = "none") +
  theme(legend.position = "bottom",
        text = element_text(size = 18)) +
  guides(color = guide_colorbar(title.position = "top")) +
  scale_size_area(max_size = 5) +
  theme(legend.text = element_text(size = 10),
        legend.title = element_text(size = 12)) +
  labs(color = "Total downloads (%)")
```

@@

Figure `r ref("fig:ctv-summ-plot")` shows that the `r ctv("ExperimentalDesign")` task view has the lowest average number of contributors among all `r n_distinct(ctv_gini_yearly$name)` CRAN task views. In addition, we can also see in Figure `r ref("fig:ctv-summ-plot")`  that the `r ctv("ExperimentalDesign")` task view has one of the least intra-connectivity (the percentage of packages that make use of other packages within the same task view). The full connection of DoE packages with other DoE packages is shown in Figure `r ref("fig:plot-doe-network")`. These observations suggest that the **_field of experimental design is one of the least collaborative field_** and package development generally occur in silos.

```{r ctv-summ-plot, fig.width = 5.5, fig.height = 6, fig.cap = "The above figure is a scatterplot of the intra-connectivity (the percentage of packages that depends, suggest or imports at least one other package within the same task view) and the average number of contributors for each CRAN task view. A low intra-connectivity suggests that development within the topic mostly occur in silos whilst high  intra-connectivity suggests that there are more interactions within the topic. The color shows the number of packages, the size of the point corresponds to the total number of contributors, and the text labels show the CRAN task view name.  The label of ExperimentalDesign task view is colored red. The task views in the bottom-left corner are topics that are more indicative of contributors working in silos. The actual numerical values are show in Table S1 in the Supplementary Material."}
ctv_summary_table %>% 
  ggplot(aes(avg_contributors, intra_perc)) +
  geom_point(aes(size = ncontributors,
                 color = n)) +
  ggnewscale::new_scale_color() + 
  ggrepel::geom_text_repel(aes(label = name,
                               color = name=="ExperimentalDesign"),
                           force = 150,
                           max.iter = 20000) +
  guides(color = "none") + 
  scale_color_manual(values = c("black", "#CC3311")) + 
  labs(x = "Average number of contributors",
       y = "Intra-connectivity (%)",
       size = "Total number of contributors",
       color_new = "Total number of packages") +
  theme(legend.position = "bottom",
        legend.box = "vertical",
        legend.box.just = "left")
```
